 

### 2021-05-19 vertical-thermal-review.Rmd
Including Song et al. 2020

This balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994] such that, because stomatal conductance ($g_s$) facilitates heat dissipation, when water demand exceeds water transport, driven by irradiance,$T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.

@@

@@

Across forest species, widely, show fundamental trait adaptation to light gradients similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. However, between multi-species canopy and understory, trait plasticity to similar biophysical gradients can vary several-folds due to inherent specie-specific traits. Spatio-temporal gradients offer another axis for trait variation, as they are heterogenous in time and space [Ishii and Asano, 2007]. Across ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. Therefore, while general patterns of thermoregulation can be observed across species, some might be more suspectable to heat stress at a vertical gradient than others [@fauset_differences_2018].

@@

While in a few studies $Q_{10}$ across species shows to be greater in upper-canopy leaves than in shaded lower-canopy leaves, with differing values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996]. Contranstingly, in a lowland tropical Australian forest study, $Q_{10}$ did not vary across species vertical profiles [@weerasingheCanopyPositionAffects2014], similary across temperate trees at Coweeta, though with a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased (*see their Fig. 1*)[@bolstad_foliar_1999]. (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*). @bolstad_foliar_1999 also found evidence of acclimation, but there remained a declining trend of respiration with elevation.

>>>>

Stomatal conductance ($g_s$) is a key factor mediating $T_{leaf}$ [@songTreeSurfaceTemperature2020] because leaf energy balance is a tradeoff between water use strategy and optimum $T_{leaf}$ [@fauset_differences_2018; @kochDiurnalPatternsLeaf1994] such that, because $gs$ facilitates heat dissipation, when water demand exceeds water transport, driven by for example: irradiance,$T_{air}$, $VPD$ or drought, conservation of water ($gs$ limitation) occurs at the cost of increasing $T_{leaf}$.

@@

Lastly, $gs$ and leaf size together are most important factors in deciding $T_{leaf}$ - $T_{air}$ and temperature sensitivity, because $gs$ determines leaf biological-biophysical interaction and leaf size determines leaf heating and cooling rate. Therefore, smaller leaves with greater $gs$ have are less sensitive to high temperature [@songTreeSurfaceTemperature2020].

@@

Across forest species, widely, show fundamental trait adaptation to light gradients similar to within-canopy traits [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. However, between multi-species canopy and understory, trait plasticity to similar biophysical gradients can vary several-folds due to inherent specie-specific traits. Spatio-temporal gradients offer another axis for trait variation, as they are heterogenous in time and space [Ishii and Asano, 2007]. Across ontogeny of the same species in canopy and understory, traits such as LMA, photosynthetic capacity [per area] and $gs$ increase with height, separate from light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012]. Species with smaller leaves and larger $gs$ will be less vulnerable to temperature increase [@fauset_differences_2018; @songTreeSurfaceTemperature2020].

@@

While in a few studies $Q_{10}$ shows to be greater in upper-canopy leaves than in shaded lower-canopy leaves across species, with differing values between species [@turnbullScalingFoliarRespiration2003; @harleyEffectsLightTemperature1996]. Contranstingly, in a lowland tropical Australian forest study, $Q_{10}$ did not vary across species vertical profiles [@weerasingheCanopyPositionAffects2014], similary across temperate trees at Coweeta, though with a slight tendency to increase down the canopy relative to upper-canopy leaves as $T_{air}$ increased (*see their Fig. 1*)[@bolstad_foliar_1999]. (*Martijn has some unpublished data on this that we might include.unpublished data showing higher Q10 in shade than in sun leaves for 10 species or so. May make sense to include here: https://github.com/EcoClimLab/vertical-thermal-review/issues/8.*). @bolstad_foliar_1999 also found evidence of acclimation, but there remained a declining trend of respiration with elevation.
- Response of respiration to leaf temperature [partially represented  by $E_{0}$] in Quercus rubra hardly varies with canopy position, but canopy-position effect is respiration is much greater in mesic sites, and seasonal effect is greater in drier sites [@xuSeasonalVariationTemperature2006]
